---
title: "ClassifyCurves"
author: "Dan Weinberger"
date: "November 12, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Classify growth curves based on trends following exponential growth phase
These growth data were collected by Adrienn Tothpal and colleagues as described in https://doi.org/10.1101/416040 . Pneumococci were grown in BHI broth in microtiter plates under a range of temperature and oxygen levels. Functional Data Analysis (FDA) is a method for extracting key information from time series data and summarizing the curves using a series of functions. This allows for easy calculation of derivatives and the decomposition of the curves into principal components based onvariability in the functions. With FPCA, smooth, flexible functions are fit through the data, and for each function, each growth curve is assigned a multiplier (PC). By multiplying the multiplier by the function and adding the functions to the mean trajectory, it is possible to recreate the original curve. For more info on FPCA, see Principal Modes of Variation for Processes With Continuous Sample Curves Castro, Lawton, SYLVESTRE and a strightforwrd review: https://www.tandfonline.com/doi/full/10.1080/14763141.2017.1392594?src=recsys

```{r loadpkg, include=FALSE}
packages = c("grofit","reshape","RColorBrewer","dplyr",'fdapace', 'EMCluster','aplpack','ks','nlme' )
#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
    }
})
#search()

```

```{r pressure, echo=FALSE}
##IMPORT AND FORMAT DATA
library(RCurl)
x<-getURL('https://raw.githubusercontent.com/weinbergerlab/GrowthVariation/master/master%205_24_2018.csv')
d1a<- read.csv(text=x)
d1a$st[which(d1a$st=="11")]<-"11A"
d1a$st[which(d1a$st=="15")]<-"15B"
d1a$st[d1a$st=='5' &d1a$Diagnosis==5]<-"14"  #Our ST 5 strain from CDC is actually ST 14
#d1a$dx<-NA
#d1a$dx[d1a$Diagnosis ==0]<-0
#d1a$dx[d1a$Diagnosis %in% c(5)]<-5
#d1a$dx[d1a$Diagnosis %in% c(2)]<-2
d1a$st=as.character(d1a$st)
d1a$Diagnosis=as.numeric(d1a$Diagnosis)
d1a$X<-NULL
d1a$Age<-NULL
d1a$Agegroup=NULL
d1a$Ery<-NULL
d1a$Pen<-NULL
d1a$Pathogen<-NULL
d1a$Pilus<-NULL
d1a$Variant<-NULL
#index<-1:nrow(d1a)
ID2<-paste(d1a$Group,d1a$st,d1a$Diagnosis,sep="_")
d1a<-cbind.data.frame(ID2, d1a)
d1a$ID<-as.character(d1a$ID)
d1a$ID[d1a$ID==""]<-as.character(d1a$ID2[d1a$ID==""])
d1a$ID<-as.factor(d1a$ID)
d1a$Group<-NULL
d1a$Diagnosis[substr(d1a$st,1,3)=="603"]<-7 #isogenic
d1a$Diagnosis[substr(d1a$st,1,4)=="TIGR"]<-8 #isogenic
d1a$Diagnosis[d1a$st=="15B-" | d1a$st=='10A-']<-9 #cps- knockout
d1a<-d1a[substr(d1a$st,1,1) != "P",]
d1a<-d1a[d1a$st != "SAUR",]
d1a<-d1a[d1a$st != "Morax",]
d1a<-d1a[d1a$st != "Blank",]
d1a<-d1a[d1a$st != "",]
d1a$st[d1a$st %in% c("15B", "15C", '15B/C')] <-"15BC"
#unique(d1a$st)
## blank based on t=0
d1a$X1[d1a$X1>d1a$X2]<- d1a$X2[d1a$X1>d1a$X2]
d1a[,7:ncol(d1a)]<-d1a[,7:ncol(d1a)] - (d1a$X1) #
d1a[,7:ncol(d1a)][d1a[,7:ncol(d1a)]<0]<-0  #Values must be >=0
d1a<-d1a[d1a$temp>=30 & d1a$temp<=39,]
d1a<-d1a[d1a$st !="unknown",]
d1a<-d1a[d1a$st !="GBS",]
d1a<-d1a[which(!is.na(d1a$Diagnosis)),]
d1a$st[d1a$st=="NT"]<-"cps-"
d1a$st[d1a$st=="TIGR4 Janus"]<-"cps-"
d1a$st[d1a$st=="603 Janus"]<-"cps-"
d1a$st[d1a$st=="603"]<-"6B"
d1a$st[d1a$st=="TIGR4"]<-"4"
d1a$st[d1a$st=="TIGR5"]<-"5"
d1a$st[d1a$st=="TIGR14"]<-"14"
d1a$st[d1a$st=="TIGR19F"]<-"19F"
d1a$st[d1a$st=="R6"]<-"cps-"
d1a<-d1a[d1a$temp>=33 & d1a$temp<=37 &d1a$anaerobic %in% c('1','2'),] #Restrict to 30-39C temperatures (readings at lower temps not reliable)

gromax<-apply(d1a[,7:ncol(d1a)],1, max, na.rm=TRUE)
d1a<-d1a[gromax>=0.1,] #if no growth above OD600=0.1, filter out the row

```

##Fit spline to data, calculate derivative
Use smooth.spline function, which is same thing they do in the grofit package; use a fixed smoothing parameter for simplicity
```{r prep_spline}
Y.list<- split(d1a[,-c(1:8, (48+8+1):ncol(d1a))], as.factor(1:nrow(d1a)))
d1.data<-d1a[,-c(1:8, (48+8+1):ncol(d1a))]
T.list<-lapply(Y.list, function(x) (1:length(x))/2 ) #List of times (h)


  spl1<-lapply(Y.list, function(x) smooth.spline(x[!is.na(x)], spar=0.5))
  fit1<-t(sapply(spl1, function(x) c(predict(x)$y, rep(NA, 48-length(predict(x)$y ))  )))
  deriv1<-t(sapply(spl1, function(x) c(predict(x, deriv=1)$y, rep(NA, 48-length(predict(x)$y ))  )))
  deriv2<-t(sapply(spl1, function(x) c(predict(x, deriv=2)$y, rep(NA, 48-length(predict(x)$y ))  )))
                              
```


```{r extractderiv , echo=FALSE}
#Extract information on derivatives (maximum and minimum derivatives, timing of max and min)
max.deriv1<-apply(deriv1,1,function(x) max(x, na.rm=TRUE))
max.deriv2<-apply(deriv2,1,function(x) max(x, na.rm=TRUE))
deriv1.start.increase<-apply(deriv1,1,function(x){
  which(lag(x)<0 & x>0)
} )

#What is time of max 1st derivative?
#prior to the max slope, set to NA so that cna calculate minimum deriv AFTER the maximum
t.max.deriv1<-rep(NA, length=length(max.deriv1))
derivs.all.post.max<-deriv1
for(i in 1: nrow(deriv1)){
  t.max.deriv1[i]<-which(deriv1[i,]==max.deriv1[i])[1]
  derivs.all.post.max[i,1:(t.max.deriv1[i]-1)]<-NA
}

min.deriv1<-apply(derivs.all.post.max,1,function(x) min(x, na.rm=TRUE))
min.deriv2<-apply(derivs.all.post.max,1,function(x) min(x, na.rm=TRUE))
min.deriv1.0<-min.deriv1
#min.deriv1.0[min.deriv1>0]<-0 #if never declines, set minimum of deriv1 to 0
```

#Classify curves based on deriatives
With first derivative: if density increases then decreases, should have large positive value followed by large negative value. If fensity increases then plateau, derivative will be positive then move towards 0; if inceases then increases again will be positive, then decrease, then increase again. Take difference between max and min and scale by max. If large max and plateau, this value will be close to 1; if large max followed by equally large min, value will be ~0.5

Simple classifier:  (max.deriv-min.deriv)/ Max(deriv1). If this value is >1, then decline is faster than then the max exponential increase; if decline  e of decrease equals max rate of increase.  closer to 1 indicates flat

Only consider curves that have max growth rate earlier than 10 hours. Otherwise, can't evaluate properly whether it ultimately declines.
```{r curve.classify}
#Ratio of max rate of increase vs max rate of decrease if decreases
trans.black<-rgb(0,0,0, alpha=0.3)

ratio_min_max_rate<- (max.deriv1-min.deriv1.0)/max.deriv1 #smaller values=faster decline relative to increase 
hist(ratio_min_max_rate)
curve.cuts<-list(c(0,1.1),c(1.1,2.0),c(2.0 ,max(ratio_min_max_rate)))
par(mfrow=c(length(curve.cuts),4),mar=c(1,1,1,1))
for(i in 1:length(curve.cuts)){
increase.indices<-which(ratio_min_max_rate<curve.cuts[[i]][2] & ratio_min_max_rate>=curve.cuts[[i]][1] & t.max.deriv1<=15) #What is cutoff--lower values give ones that decrease less; 
increase.indices.plot<-sample(increase.indices, size=min(50, length(increase.indices)))
matplot(t(d1.data[increase.indices.plot,]), type='l', lty=1, bty='l')
title(paste0("Observed ",i))
matplot(t(fit1[increase.indices.plot,]), type='l', lty=1, bty='l')
title(paste0("Smooth Observed ",i))
matplot(t(deriv1[increase.indices.plot,]), type='l', lty=1, bty='l')
abline(h=0, col='red')
title("1st Derivative")
matplot(t(deriv2[increase.indices.plot,]), type='l', lty=1, bty='l')
abline(h=0, col='red')
title("2nd Derivative")
}
flatest.grp<-which(ratio_min_max_rate<curve.cuts[[1]][2] & ratio_min_max_rate>=curve.cuts[[1]][1] & t.max.deriv1<=25) 
increasing.samples<-d1a[flatest.grp, c(1:6)]
increasing.samples<-increasing.samples[order(increasing.samples$st),]
increasing.samples
```
#Can we pull out specifically the curves that have 2 exponential growth phases?
Should show up as a double hump on the 1st derivative. 11A, 3, 8, 18A pop out on this analysis; 3 and 8 have almost identical motifs (Glc-GlcA). 11A and 18A both have Glycerol

```{r curve.classify2}
#Ratio of max rate of increase vs max rate of decrease if decreases
trans.black<-rgb(0,0,0, alpha=0.3)

ratio_min_max_rate<- (max.deriv1-min.deriv1.0)/max.deriv1 #smaller values=faster decline relative to increase 
hist(ratio_min_max_rate)
curve.cuts<-list(c(0,1.0),c(1.0,1.5),c(1.5,2.0),c(2.0 ,max(ratio_min_max_rate)))
par(mfrow=c(length(curve.cuts),4),mar=c(1,1,1,1))
for(i in 1:length(curve.cuts)){
increase.indices<-which(ratio_min_max_rate<curve.cuts[[i]][2] & ratio_min_max_rate>=curve.cuts[[i]][1] & t.max.deriv1<=25) #What is cutoff--lower values give ones that decrease less; 
increase.indices.plot<-sample(increase.indices, size=min(50, length(increase.indices)))
matplot(t(d1.data[increase.indices.plot,]), type='l', lty=1, bty='l')
title(paste0("Observed ",i))
matplot(t(fit1[increase.indices.plot,]), type='l', lty=1, bty='l')
title(paste0("Smooth Observed ",i))
matplot(t(deriv1[increase.indices.plot,]), type='l', lty=1, bty='l')
abline(h=0, col='red')
title("1st Derivative")
matplot(t(deriv2[increase.indices.plot,]), type='l', lty=1, bty='l')
abline(h=0, col='red')
title("2nd Derivative")
}
flatest.grp<-which(ratio_min_max_rate<curve.cuts[[1]][2] & ratio_min_max_rate>=curve.cuts[[1]][1] & t.max.deriv1<=30) 
increasing.samples<-d1a[flatest.grp, c(1:6)]
increasing.samples<-increasing.samples[order(increasing.samples$st),]
increasing.samples
```



#Evaluate characteristics of curves that have smaller or larger values based on our simple classifier
```{r correlate_flatness}
mod.df<-cbind.data.frame(ratio_min_max_rate=ratio_min_max_rate,id=d1a$ID, st=d1a$st, anaerobic=as.factor(d1a$anaerobic),Diagnosis=as.factor(d1a$Diagnosis), t.max.deriv1=t.max.deriv1, temp= as.factor(d1a$temp))
mod1<-lme(ratio_min_max_rate ~ st+  t.max.deriv1 + temp*anaerobic , random=~1|id, data=mod.df[mod.df$anaerobic !='0',] )  
round(summary(mod1)$tTable,3)
coef1<-mod1$coefficients$fixed
coef.st<-coef1[substr(names(coef1),1,2)=='st']
#sort(coef.st)
print(sort(coef.st))
```

Does the pattern differ between IPD and carriage isolates (dx=5 vs dx=0)? No evidence that it does. 
```{r dx.effect}
mod1<-lm(ratio_min_max_rate ~ st+  t.max.deriv1 + temp*anaerobic +Diagnosis, data=mod.df[mod.df$anaerobic !='0' & mod.df$temp %in% c('33','35','37'), ] )  
summary(mod1)
```


Evaluating the interaction between temperature and oxygen; we can see that curves are flatter in aerobic conditions with catalase. (bigger values=greater declines)

```{r flatness.int}
  interaction.plot(mod.df$temp, mod.df$anaerobic, mod.df$ratio_min_max_rate, bty='l')
   title('Interaction of temperature and oxygen')
   
   ps2.mid<-mod.df[mod.df$temp %in% c('33','35','37') & mod.df$anaerobic %in% c('1','2'),]
     interaction.plot(ps2.mid$anaerobic,ps2.mid$st,  ps2.mid$ratio_min_max_rate, bty='l')
   title('Interaction of ST and oxygen')
```

Same thing but restrict to 33-37C and stratify by aerobic vs anaerobic

```{r correlate_flatness2}
mod.df<-cbind.data.frame(ratio_min_max_rate=ratio_min_max_rate,id=d1a$ID, st=d1a$st, anaerobic=as.factor(d1a$anaerobic), t.max.deriv1=t.max.deriv1, temp= as.factor(d1a$temp))
mod2<-lme(ratio_min_max_rate ~ st+ t.max.deriv1 + temp*anaerobic , random=~1|id, data=mod.df[mod.df$anaerobic !='0' & mod.df$temp %in% c('33','35','37'),] )  
#round(summary(mod2)$tTable,3)
coef2<-mod2$coefficients$fixed
coef.st2<-coef2[substr(names(coef2),1,2)=='st']
#sort(coef.st)
print(sort(coef.st2))
```

Aerobic (catalase) only
```{r correlate_flatness2a}
mod4<-lme(ratio_min_max_rate ~ st+ t.max.deriv1 + temp  , random=~1|id, data=mod.df[mod.df$anaerobic %in% c('2') & mod.df$temp %in% c('33','35','37'),] ) 
#round(summary(mod4)$tTable,3)
coef4<-mod4$coefficients$fixed
coef.st4<-coef4[substr(names(coef4),1,2)=='st']
print(sort(coef.st4))
```
Anaerobic only
```{r correlate_flatness2b}
mod4<-lme(ratio_min_max_rate ~ st+ t.max.deriv1 + temp  , random=~1|id, data=mod.df[mod.df$anaerobic %in% c('1') & mod.df$temp %in% c('33','35','37'),] ) 
#round(summary(mod4)$tTable,3)
coef4<-mod4$coefficients$fixed
coef.st4<-coef4[substr(names(coef4),1,2)=='st']
print(sort(coef.st4))
```

```{r covar_setup, echo=FALSE}
#Covariate data
x<-getURL('https://raw.githubusercontent.com/weinbergerlab/GrowthVariation/master/pneumo_master%20v2.csv')
cov1<-read.csv(text=x)
cov1$st<-as.character(cov1$ST)
cov1$MASSCARR01[is.na(cov1$MASSCARR01)]<-0
cov1$OSMANCARRN[is.na(cov1$OSMANCARRN)]<-0
cov1$SLEEMANCARRN[is.na(cov1$SLEEMANCARRN)]<-0
cov1$NORWAYCARR_PRE[is.na(cov1$NORWAYCARR_PRE)]<-0
cov1$TROTTERCARRN[is.na(cov1$TROTTERCARRN)]<-0
cov1$GREECECARRN[is.na(cov1$GREECECARRN)]<-0
cov1$ave_carr_pre<-apply( cbind(cov1$MASSCARR01, cov1$OSMANCARRN, cov1$SLEEMANCARRN, cov1$NORWAYCARR_PRE, cov1$TROTTERCARRN, cov1$GREECECARRN), 1, mean, na.rm=TRUE )
cov1$st[cov1$st=='6A/C']<-'6A'
carr.comp<-merge(mod.df,cov1, by='st' )
carr.comp2<-carr.comp[carr.comp$anaerobic %in% c('1','2') &carr.comp$temp %in% c('33','35','37') & !is.na(carr.comp$ratio_min_max_rate) ,]

#CFR Harboe
  ds<-carr.comp2[,c('ratio_min_max_rate','anaerobic','t.max.deriv1', 'temp','id' )]
  ds<-ds[complete.cases(ds),]
```
#What about CPS knockout strains vs parents?
#WHat about CPS switch variants?

Tigr4 CPS switch and KO variants; doesn't show clear difference. ST14 might have more of a plateau before declining. But ratio of min max deriv is same for 14 and CPS-; nand bigger declines for TIGR5 and TIGR19F

```{r tigr, echo=FALSE}
tigr1<-d1a[d1a$Diagnosis==8 &d1a$temp==33 & d1a$anaerobic==2 ,]
matplot(t(tigr1[,7:40]), bty='l', type='l', col=c('black','red','blue','green'), lty=c(1,2,3,4))
legend(0,0.3, legend=tigr1$st,col=c('black','red','blue','green'), lty=c(1,2,3,4))

cbind(tigr1$st,ratio_min_max_rate[which(d1a$Diagnosis==8 &d1a$temp==33 & d1a$anaerobic==2 )]
)
```

#Cps knockouts

10A and 10A KO (CDC strain); Doesn't show clear effect of removing capsule. All strains have nice plateau

```{r ko10A, echo=FALSE}
ko10A.select<-which(d1a$Diagnosis==9 & d1a$temp==33 & d1a$anaerobic==2 & d1a$st=='10A-')
cps10A.select<-which(d1a$Diagnosis==5 & d1a$temp==33 & d1a$anaerobic==2 &d1a$st=='10A')

matplot(t(d1a[c(ko10A.select,cps10A.select),7:40]), bty='l', type='l', col=c('black','red','red'), lty=c(1,2))
legend(0,0.3, legend=c('CPS-','10A', '10A'),col=c('black','red', 'red'), lty=c(1,2))
```

15B and 15B KO (CDC 15B strain); No clear effect of removing capsule on plateau; CPS- seems to reach lower OD but has mre of a plateau (no decline)
```{r ko15B, echo=FALSE}
ko15B.select<-which(d1a$Diagnosis==9 & d1a$temp==33 & d1a$anaerobic==2 & d1a$st=='15B-')
cps15B.select<-which(d1a$Diagnosis==5 & d1a$temp==33 & d1a$anaerobic==2 & d1a$ID=='CDC_15B')

matplot(t(d1a[c(ko15B.select,cps15B.select),7:40]), bty='l', type='l', col=c('black','red','red'), lty=c(1,2))
legend(0,0.3, legend=c('CPS-','15B', '15B'),col=c('black','red', 'red'), lty=c(1,2))
```

6B and 6B KO strain (603/603:Janus); No notable effect f knocking out the capsule on growth pattern
```{r ko6B, echo=FALSE}
ko6B.select<-which(d1a$Diagnosis==7 & d1a$temp==33 & d1a$anaerobic==2 )


matplot(t(d1a[c(ko6B.select),7:40]), bty='l', type='l', col=c('red','black'), lty=c(1,2))
legend(0,0.3, legend=c('6B','CPS-'),col=c('red','black'), lty=c(1,2))
```


##Analyze association between growth curve pattern and epi/Ps characteristics
These analyses are basically the same as regression above, but drop serotype fixed effect and instead include serotype characteristics. This shows correlation with Ps complexity and the presence of uronic acids (more complex Ps, bigger crash;more uronic acid, smaller crash) 

#Case fatality ratio (Harboe)
```{r epi_cor}
carr.comp2<-carr.comp[carr.comp$anaerobic %in% c('1','2') &carr.comp$temp %in% c('33','35','37') & !is.na(carr.comp$ratio_min_max_rate) ,]

#CFR Harboe
  ds<-carr.comp2[,c('ratio_min_max_rate', 'CFRHARBOE','anaerobic','t.max.deriv1', 'temp','id' )]
  ds<-ds[complete.cases(ds),]
  reg.cfr<-lme(ratio_min_max_rate ~ CFRHARBOE + anaerobic +t.max.deriv1 + temp , random=~1|id, data=ds )  
  round(summary(reg.cfr)$tTable,3)
   plot(carr.comp2$ratio_min_max_rate, carr.comp2$CFRHARBOE, col='white')
   text(carr.comp2$ratio_min_max_rate, carr.comp2$CFRHARBOE, carr.comp2$st)
```

Polysaccharide complexity
```{r epi_cor2}
carr.comp2<-carr.comp[carr.comp$anaerobic %in% c('1','2') &carr.comp$temp %in% c('33','35','37') & !is.na(carr.comp$ratio_min_max_rate) ,]

 #Total carbon
  ds<-carr.comp2[,c('ratio_min_max_rate', 'TOTALCARBONREPEAT','anaerobic','t.max.deriv1', 'temp','id' )]
  ds<-ds[complete.cases(ds),]
  reg.carbon<-lme(ratio_min_max_rate ~ TOTALCARBONREPEAT + anaerobic +t.max.deriv1 + temp , random=~1|id, data=ds )
  round(summary(reg.carbon)$tTable,3)
   plot(carr.comp2$ratio_min_max_rate, carr.comp2$TOTALCARBONREPEAT, col='white')
   text(carr.comp2$ratio_min_max_rate, carr.comp2$TOTALCARBONREPEAT, carr.comp2$st)
```

Invasiveness (AJE paper)--inverse variance weights(note regression not weighted)
```{r epi_cor3}
carr.comp2<-carr.comp[carr.comp$anaerobic %in% c('1','2') &carr.comp$temp %in% c('33','35','37') & !is.na(carr.comp$ratio_min_max_rate) ,]
  x<-getURL('https://raw.githubusercontent.com/weinbergerlab/GrowthVariation/master/mcmc_invasive_single_stage.csv')
   inv1<-read.csv(text=x)
   inv1$st=as.character(inv1$st)
   inv1$st[inv1$st=='6A/C']<-'6A'
   inv2<-merge(carr.comp, inv1, by='st' , all=TRUE)
    ds<-inv2[,c('ratio_min_max_rate', 'log.inv.prec.age1','log.inv.age1','anaerobic','t.max.deriv1', 'temp','id' )]
   ds<-ds[complete.cases(ds),]
   trans.red<-rgb(1,0,0, alpha=0.1)
    symbols(sqrt(ds$ratio_min_max_rate), ds$log.inv.age1, sqrt(ds$log.inv.prec.age1/pi),inches=0.35,     
            bty='l',fg="white", bg=trans.red,         xlab="Ratio increasing/decreasing Rates", ylab="log(Invasiveness")
   text(sqrt(ds$ratio_min_max_rate), inv2$log.inv.age1, inv2$st, cex=0.75)
  reg.inv<-lme(ratio_min_max_rate ~ log.inv.age1 + anaerobic +t.max.deriv1 + temp , random=~1|id, data=ds )
  round(summary(reg.inv)$tTable,3)
```

Specific polysaccharide components (GlcA/GalA or NAc sugars)
```{r epi_cor4}
carr.comp2<-carr.comp[carr.comp$anaerobic %in% c('1','2') &carr.comp$temp %in% c('33','35','37') & !is.na(carr.comp$ratio_min_max_rate) ,]
   ps1<-read.csv('PS Composition_SS_final.csv')
   ps1$st<-as.character(ps1$Serotype)
   ps2<-merge(carr.comp2, ps1, by='st' , all=TRUE)
   ps2$NAC<-0
   ps2$NAC[ps2$GlcNAc==1 | ps2$GalNAc==1 | ps2$ManNAc==1 |  ps2$ManNAcA==1 |ps2$FucNAc==1 |ps2$PneNAc ==1] <-1
   ps2$uronic<-0
   ps2$uronic[ps2$GalA==1 |ps2$GlcA==1 ]<-1  #GlcA and GalA derived from same pathway
   plot(as.factor(ps2$uronic), ps2$ratio_min_max_rate)
   plot(as.factor(ps2$uronic[ps2$anaerobic=='1']), ps2$ratio_min_max_rate[ps2$anaerobic=='1'])
   plot(as.factor(ps2$uronic[ps2$anaerobic=='2']), ps2$ratio_min_max_rate[ps2$anaerobic=='2'])
   plot(as.factor(ps2$Gro[ps2$anaerobic=='2']), ps2$ratio_min_max_rate[ps2$anaerobic=='2'])
   
      ds<-ps2[ps2$anaerobic!='0',c('ratio_min_max_rate', 'uronic','NAC','Gro','anaerobic','t.max.deriv1', 'temp','id')]
   ds<-ds[complete.cases(ds),]
   reg.ps<-lme(ratio_min_max_rate ~ uronic +NAC +Gro+ anaerobic +t.max.deriv1 + temp + anaerobic*temp, random=~1|id, data=ds )
     round(summary(reg.ps)$tTable,3)
```


#Does association between CPS component and growth vary by temp?
In Adrienn's experiments, she found that adding purified CPS had a big effect at 37, no effect or even inhibitory at 33C. Here it appears affect of uronic strongest at 33C with anaerobic; weaker at 33C aerobic 37 anaerobic in par with 33 aerobic; 37 anaerobic no effect

```{r effect_temp}

 reg.ps1<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='33' & ds$anaerobic=='1',] )
     round(summary(reg.ps1)$tTable,3)

 reg.ps2<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='33' & ds$anaerobic=='2',] )
      round(summary(reg.ps2)$tTable,3)

  reg.ps3<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='37' & ds$anaerobic=='1',] )
     round(summary(reg.ps3)$tTable,3)

 reg.ps4<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='37' & ds$anaerobic=='2',] )
     round(summary(reg.ps4)$tTable,3)
     
          #Interaction
reg.ps.int<-lme(ratio_min_max_rate ~ uronic*anaerobic +NAC*anaerobic + anaerobic +t.max.deriv1 + temp + anaerobic*temp, random=~1|id, data=ds )
     round(summary(reg.ps.int)$tTable,3)

```
