---
title: "fpca"
author: "Dan Weinberger"
date: "November 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

##Functional Data Analysis of growth curve data
These growth data were collected by Adrienn Tothpal and colleagues as described in https://doi.org/10.1101/416040 . Pneumococci were grown in BHI broth in microtiter plates under a range of temperature and oxygen levels. Functional Data Analysis (FDA) is a method for extracting key information from time series data and summarizing the curves using a series of functions. This allows for easy calculation of derivatives and the decomposition of the curves into principal components based onvariability in the functions. With FPCA, smooth, flexible functions are fit through the data, and for each function, each growth curve is assigned a multiplier (PC). By multiplying the multiplier by the function and adding the functions to the mean trajectory, it is possible to recreate the original curve. For more info on FPCA, see Principal Modes of Variation for Processes With Continuous Sample Curves Castro, Lawton, SYLVESTRE and a strightforwrd review: https://www.tandfonline.com/doi/full/10.1080/14763141.2017.1392594?src=recsys

```{r loadpkg, include=FALSE}
packages = c("grofit","reshape","RColorBrewer","dplyr",'fdapace', 'EMCluster','aplpack','ks','nlme' )
#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
    }
})
#search()

```

## Format the data
The data need to be cleaned and formatted to facilitate analysis

```{r pressure, echo=FALSE}
##IMPORT AND FORMAT DATA
d1a<- read.csv("master 5_24_2018.csv")
d1a$st[which(d1a$st=="11")]<-"11A"
d1a$st[which(d1a$st=="15")]<-"15B"
d1a$st[d1a$st=='5' &d1a$Diagnosis==5]<-"14"  #Our ST 5 strain from CDC is actually ST 14
#d1a$dx<-NA
#d1a$dx[d1a$Diagnosis ==0]<-0
#d1a$dx[d1a$Diagnosis %in% c(5)]<-5
#d1a$dx[d1a$Diagnosis %in% c(2)]<-2
d1a$st=as.character(d1a$st)
d1a$Diagnosis=as.numeric(d1a$Diagnosis)
d1a$X<-NULL
d1a$Age<-NULL
d1a$Agegroup=NULL
d1a$Ery<-NULL
d1a$Pen<-NULL
d1a$Pathogen<-NULL
d1a$Pilus<-NULL
d1a$Variant<-NULL
#index<-1:nrow(d1a)
ID2<-paste(d1a$Group,d1a$st,d1a$Diagnosis,sep="_")
d1a<-cbind.data.frame(ID2, d1a)
d1a$ID<-as.character(d1a$ID)
d1a$ID[d1a$ID==""]<-as.character(d1a$ID2[d1a$ID==""])
d1a$ID<-as.factor(d1a$ID)
d1a$Group<-NULL
d1a$Diagnosis[substr(d1a$st,1,3)=="603"]<-7 #isogenic
d1a$Diagnosis[substr(d1a$st,1,4)=="TIGR"]<-8 #isogenic
d1a$Diagnosis[d1a$st=="15B-" | d1a$st=='10A-']<-9 #cps- knockout
d1a<-d1a[substr(d1a$st,1,1) != "P",]
d1a<-d1a[d1a$st != "SAUR",]
d1a<-d1a[d1a$st != "Morax",]
d1a<-d1a[d1a$st != "Blank",]
d1a<-d1a[d1a$st != "",]
d1a$st[d1a$st %in% c("15B", "15C", '15B/C')] <-"15BC"
unique(d1a$st)
## blank based on t=0
d1a$X1[d1a$X1>d1a$X2]<- d1a$X2[d1a$X1>d1a$X2]
d1a[,7:ncol(d1a)]<-d1a[,7:ncol(d1a)] - (d1a$X1) #
d1a[,7:ncol(d1a)][d1a[,7:ncol(d1a)]<0]<-0  #Values must be >=0
d1a<-d1a[d1a$temp>=30 & d1a$temp<=39,]
d1a<-d1a[d1a$st !="unknown",]
d1a<-d1a[d1a$st !="GBS",]
d1a<-d1a[which(!is.na(d1a$Diagnosis)),]
d1a$st[d1a$st=="NT"]<-"cps-"
d1a$st[d1a$st=="TIGR4 Janus"]<-"cps-"
d1a$st[d1a$st=="603 Janus"]<-"cps-"
d1a$st[d1a$st=="603"]<-"6B"
d1a$st[d1a$st=="TIGR4"]<-"4"
d1a$st[d1a$st=="TIGR5"]<-"5"
d1a$st[d1a$st=="TIGR14"]<-"14"
d1a$st[d1a$st=="TIGR19F"]<-"19F"
d1a$st[d1a$st=="R6"]<-"cps-"
d1a<-d1a[d1a$temp>=30 & d1a$temp<=39,] #Restrict to 30-39C temperatures (readings at lower temps not reliable)

```

##Simpler; just fit spline to data, calculate derivative

#Prepare data for FPCA analysis
Extract the fitted values and 1st and 2nd derivatives using fitted() function

```{r prepFPCA}
Y.list<- split(d1a[,-c(1:8, (48+8+1):ncol(d1a))], as.factor(1:nrow(d1a)))
d1.data<-d1a[,-c(1:8, (48+8+1):ncol(d1a))]
Y.list<-lapply(Y.list, function(x) as.numeric(x[1,])) #List of OD600 values
T.list<-lapply(Y.list, function(x) (1:length(x))/2 ) #List of times (h)
fpca.growth<-FPCA(Y.list, T.list, optns= list(dataType='Dense', plot=TRUE))
pct.var.pca<-fpca.growth$lambda/sum(fpca.growth$lambda)*100 #percent of variation  associated with each component

fitted.all<-fitted(fpca.growth,K = 3,derOptns = list(p = 0,  kernelType = 'epan') ) #fitted
derivs.all<-fitted(fpca.growth,K = 3,derOptns = list(p = 1,  kernelType = 'epan') ) #1st deriv
derivs2.all<-fitted(fpca.growth,K = 3,derOptns = list(p = 2,  kernelType = 'epan') )#2nd deriv

#Extract PCs
pcs<-as.data.frame(fpca.growth$xiEst )
names(pcs)<-paste0('PC', 1:ncol(pcs))
gc.labels<-d1a[,1:8]

fpcs<-cbind.data.frame(gc.labels,pcs) #EXTRACT THE PCs
fpcs$st<-as.character(fpcs$st)

```

#Extract information on derivatives (maximum and minimum derivatives, timing of max and min)

```{r extractderiv }
max.deriv1<-apply(derivs.all[,c(1:30)],1,function(x) max(x, na.rm=TRUE))
max.deriv2<-apply(derivs2.all[,c(1:30)],1,function(x) max(x, na.rm=TRUE))
deriv1.start.increase<-apply(derivs.all[,c(1:30)],1,function(x){
  which(lag(x)<0 & x>0)
} )

#What is time of max 1st derivative?
#prior to the max slope, set to NA so that cna calculate minimum deriv AFTER the maximum
t.max.deriv1<-rep(NA, length=length(max.deriv1))
derivs.all.post.max<-derivs.all
for(i in 1: nrow(derivs.all)){
  t.max.deriv1[i]<-which(derivs.all[i,c(1:30)]==max.deriv1[i])[1]
  derivs.all.post.max[i,1:t.max.deriv1[i]]<-NA
}

min.deriv1<-apply(derivs.all.post.max[,c(1:30)],1,function(x) min(x, na.rm=TRUE))
min.deriv2<-apply(derivs.all.post.max[,c(1:30)],1,function(x) min(x, na.rm=TRUE))
min.deriv1.0<-min.deriv1
#min.deriv1.0[min.deriv1>0]<-0 #if never declines, set minimum of deriv1 to 0
```

#Classify curves based on deriatives
With first derivative: if density increases then decreases, should have large positive value followed by large negative value. If fensity increases then plateau, derivative will be positive then move towards 0; if inceases then increases again will be positive, then decrease, then increase again. Take difference between max and min and scale by max. If large max and plateau, this value will be close to 1; if large max followed by equally large min, value will be ~0.5

Simple classifier:  (max.deriv-min.deriv)/ Max(deriv1). If this value is >1, then decline is faster than then the max exponential increase; if decline  e of decrease equals max rate of increase.  closer to 1 indicates flat

Only consider curves that have max growth rate earlier than 10 hours. Otherwise, can't evaluate properly whether it ultimately declines.
```{r curve.classify}
#Ratio of max rate of increase vs max rate of decrease if decreases
  trans.black<-rgb(0,0,0, alpha=0.3)

ratio_min_max_rate<- (max.deriv1-min.deriv1.0)/max.deriv1 #smaller values=faster decline relative to increase 
hist(ratio_min_max_rate)
curve.cuts<-list(c(0,1.0),c(1.0,2.0),c(2.0 ,max(ratio_min_max_rate)))
par(mfrow=c(length(curve.cuts),4),mar=c(1,1,1,1))
for(i in 1:length(curve.cuts)){
increase.indices<-which(ratio_min_max_rate<curve.cuts[[i]][2] & ratio_min_max_rate>=curve.cuts[[i]][1] & t.max.deriv1<=15) #What is cutoff--lower values give ones that decrease less; 
increase.indices.plot<-sample(increase.indices, size=min(20, length(increase.indices)))
matplot(t(d1.data[increase.indices.plot,]), type='l', lty=1, bty='l')
title(paste0("Observed ",i))
matplot(t(fitted.all[increase.indices.plot,]), type='l', lty=1, bty='l')
title(paste0("Smooth Observed ",i))
matplot(t(derivs.all[increase.indices.plot,]), type='l', lty=1, bty='l')
abline(h=0, col='red')
title("1st Derivative")
matplot(t(derivs2.all[increase.indices.plot,]), type='l', lty=1, bty='l')
abline(h=0, col='red')
title("2nd Derivative")
}
flatest.grp<-which(ratio_min_max_rate<curve.cuts[[1]][2] & ratio_min_max_rate>=curve.cuts[[1]][1]) 
increasing.samples<-d1a[flatest.grp, c(1:6)]
increasing.samples<-increasing.samples[order(increasing.samples$st),]
increasing.samples
```

Evaluating the interaction betwene temperature and oxygen; we can see that curves are flatter in aerobic conditions. these differences are most pronounced at 33-37C and probably flattest at 33C. (bigger values=greater declines)

```{r flatness.int}
  interaction.plot(ps2$temp, ps2$anaerobic, sqrt(ps2$ratio_min_max_rate), bty='l')
   title('Interaction of temperature and oxygen')
   
   ps2.mid<-ps2[ps2$temp %in% c('33','35','37'),]
     interaction.plot(ps2.mid$st, ps2.mid$anaerobic, ps2.mid$ratio_min_max_rate, bty='l')
   title('Interaction of ST and oxygen')
```


#Evaluate characteristics of curves that have smaller or larger values based on our simple classifier
```{r correlate_flatness}
mod.df<-cbind.data.frame(ratio_min_max_rate=ratio_min_max_rate,id=d1a$ID, st=d1a$st, anaerobic=as.factor(d1a$anaerobic), t.max.deriv1=t.max.deriv1, temp= as.factor(d1a$temp))
mod1<-lme(ratio_min_max_rate ~ st+ t.max.deriv1 + temp*anaerobic , random=~1|id, data=mod.df[mod.df$anaerobic !='0',] )  
round(summary(mod1)$tTable,3)
coef1<-mod1$coefficients$fixed
coef.st<-coef1[substr(names(coef1),1,2)=='st']
#sort(coef.st)
print(sort(coef.st))
```

#Same thing but restrict to 33-37C and stratify by aerobic vs anaerobic
```{r correlate_flatness}
mod.df<-cbind.data.frame(ratio_min_max_rate=ratio_min_max_rate,id=d1a$ID, st=d1a$st, anaerobic=as.factor(d1a$anaerobic), t.max.deriv1=t.max.deriv1, temp= as.factor(d1a$temp))
mod2<-lme(ratio_min_max_rate ~ st+ t.max.deriv1 + temp*anaerobic , random=~1|id, data=mod.df[mod.df$anaerobic !='0' & mod.df$temp %in% c('33','35','37'),] )  
round(summary(mod1)$tTable,3)
coef2<-mod2$coefficients$fixed
coef.st2<-coef2[substr(names(coef2),1,2)=='st']
#sort(coef.st)
print(sort(coef.st2))

mod3<-lme(ratio_min_max_rate ~ st*anaerobic+ t.max.deriv1 + temp  , random=~1|id, data=mod.df[mod.df$anaerobic %in% c('1','2') & mod.df$temp %in% c('33','35','37'),] ) 
round(summary(mod3)$tTable,3)
coef3<-mod3$coefficients$fixed
coef.st3<-coef3[substr(names(coef3),1,2)=='st']
#print(sort(coef.st3))

```

#Visualize how the growth curves vary according to Principal Components
Indicates first set of variation is in height of curve; second is largely in timing, 3rd seems to be along whether it flattens or declines post peak

```{r plot_pca, echo=FALSE}

#####
#FUNCTION TO EXTRACT fitted value based on 1st or second PCs
# based on " Displaying the Important Features of Large Collections of Similar Curves" Jones and Rice 1989
plot_func<-function(fpcaObj, plot.quants){
  #Eigenfunctions
  eigenfunc<-fpcaObj$phi  #Eigenfunction
  xiEst2<-fpcaObj$xiEst     
  est.meanfunc<-fpcaObj$mu  #Mean function
  xi.est2.quantiles<- apply(xiEst2,2, quantile, probs=plot.quants)
  yEst.q<-array(NA, dim=c(length(est.meanfunc),3,ncol(xi.est2.quantiles) ))
  #Just reconstruct based on PCN
  for(pcn in 1:ncol(xi.est2.quantiles) ){
    yEst.q[,,pcn]<- t(xi.est2.quantiles[,pcn] %*%  t(eigenfunc[,pcn]) + t(matrix(rep(est.meanfunc,3), nrow=length(est.meanfunc))) )
  }
  return(yEst.q)
}
quants.pca<-plot_func(fpca.growth,plot.quants=c(0.1,0.5,0.9))
par(mfrow=c(2,2))
for(i in 1:4){
matplot(quants.pca[,,i], type='l', bty='l', col=c('#1b9e77','#d95f02','#7570b3' ), lty=c(2,1,3), lwd=c(1,2,1)) #Plot 1st principal component range
  title(paste0("Variation associated with PCA ", i,"; ", round(pct.var.pca[i],0),"%"))
}
```

#Visualize the variaition in PC3--plot deviations from mean for low, med, high levels (from left to right) of PC3 (note this is ignoring the variation in PC1 and PC2) On further inspection, it seems that PC3 itself is a bit misleading when interpreted without context of PC1 and PC2
```{r plot_pca3_only, echo=FALSE}
#VISUALIZE HOW GROWTH CURVES VARY ALONG PC3
plot_func<-function(fpcaObj, plot.quants){
  #Eigenfunctions
  eigenfunc<-fpcaObj$phi  #Eigenfunction
  xiEst2<-fpcaObj$xiEst     
  est.meanfunc<-fpcaObj$mu  #Mean function
  xi.est2.quantiles<- apply(xiEst2,2, quantile, probs=plot.quants)
  yEst.q<-array(NA, dim=c(length(est.meanfunc),nrow(xiEst2),ncol(xiEst2) ))
  #Just reconstruct based on PCN
  for(pcn in 1:ncol(xiEst2) ){
    yEst.q[,,pcn]<- t(xiEst2[,pcn, drop=FALSE] %*%  t(eigenfunc[,pcn]) + t(matrix(rep(est.meanfunc,nrow(xiEst2)), nrow=length(est.meanfunc))) )
  }
  return(yEst.q)
}
pca.all<-plot_func(fpca.growth,plot.quants=c(0.1,0.5,0.9))
pc3<-fpcs$PC3
pc2<-fpcs$PC2

pc2.q<-quantile(pc2, probs=c(0.05, 0.1,0.2,0.4,0.6,0.8))
pc3.q<-quantile(pc3, probs=c(0.05, 0.1,0.2,0.4,0.6,0.8))
pc3.grp1<- which(pc3<pc3.q[1])
pc3.grp2<-which(pc3>=pc3.q[1] & pc3<pc3.q[2])
pc3.grp3<-which(pc3>=pc3.q[2] & pc3<pc3.q[3])
pc3.grp4<-which(pc3>=pc3.q[3] & pc3<pc3.q[4])
pc3.grp5<-which(pc3>=pc3.q[4] & pc3<pc3.q[5])
pc3.grp6<-which(pc3>=pc3.q[5] & pc3<pc3.q[6])
pc3.grp7<-which(pc3>=pc3.q[6])
#PC2 score is high
pc2.grp6<-which(pc2>=pc2.q[5] & pc2<pc2.q[6])
pc2.grp7<-which(pc2>=pc2.q[6])

#Plot this variability
par(mfrow=c(2,4))
matplot(pca.all[,pc3.grp1,2], type='l', col=trans.black, bty='l')
matplot(pca.all[,pc3.grp2,2], type='l', col=trans.black, bty='l')
matplot(pca.all[,pc3.grp3,2], type='l', col=trans.black, bty='l')
matplot(pca.all[,pc3.grp4,2], type='l', col=trans.black, bty='l')
matplot(pca.all[,pc3.grp5,2], type='l', col=trans.black, bty='l')
matplot(pca.all[,pc3.grp6,2], type='l', col=trans.black, bty='l')
matplot(pca.all[,pc3.grp7,2], type='l', col=trans.black, bty='l')

```

#Compile covariates; generate average pre-vaccine carriage variable
```{r covar_setup}
#Covariate data
cov1<-read.csv('pneumo_master v2.csv')
cov1$st<-as.character(cov1$ST)
cov1$MASSCARR01[is.na(cov1$MASSCARR01)]<-0
cov1$OSMANCARRN[is.na(cov1$OSMANCARRN)]<-0
cov1$SLEEMANCARRN[is.na(cov1$SLEEMANCARRN)]<-0
cov1$NORWAYCARR_PRE[is.na(cov1$NORWAYCARR_PRE)]<-0
cov1$TROTTERCARRN[is.na(cov1$TROTTERCARRN)]<-0
cov1$GREECECARRN[is.na(cov1$GREECECARRN)]<-0
cov1$ave_carr_pre<-apply( cbind(cov1$MASSCARR01, cov1$OSMANCARRN, cov1$SLEEMANCARRN, cov1$NORWAYCARR_PRE, cov1$TROTTERCARRN, cov1$GREECECARRN), 1, mean, na.rm=TRUE )
cov1$st[cov1$st=='6A/C']<-'6A'
carr.comp<-merge(mod.df,cov1, by='st' )
```



#Analyze association between growth curve pattern and epi/Ps characteristics
These analyses are basically the same as regression above, but drop serotype fixed effect and instead include serotype characteristics. This shows correlation with Ps complexity and the presence of uronic acids (more complex Ps, bigger crash;more uronic acid, smaller crash) 
```{r epi_cor}
carr.comp2<-carr.comp[carr.comp$anaerobic %in% c('1','2') &carr.comp$temp %in% c('33','35','37') & !is.na(carr.comp$ratio_min_max_rate) ,]

#CFR Harboe
  ds<-carr.comp2[,c('ratio_min_max_rate', 'CFRHARBOE','anaerobic','t.max.deriv1', 'temp','id' )]
  ds<-ds[complete.cases(ds),]
  reg.cfr<-lme(ratio_min_max_rate ~ CFRHARBOE + anaerobic +t.max.deriv1 + temp , random=~1|id, data=ds )  
  round(summary(reg.cfr)$tTable,3)
   plot(carr.comp2$ratio_min_max_rate, carr.comp2$CFRHARBOE, col='white')
   text(carr.comp2$ratio_min_max_rate, carr.comp2$CFRHARBOE, carr.comp2$st)

 #Total carbon
  ds<-carr.comp2[,c('ratio_min_max_rate', 'TOTALCARBONREPEAT','anaerobic','t.max.deriv1', 'temp','id' )]
  ds<-ds[complete.cases(ds),]
  reg.carbon<-lme(ratio_min_max_rate ~ TOTALCARBONREPEAT + anaerobic +t.max.deriv1 + temp , random=~1|id, data=ds )
  round(summary(reg.carbon)$tTable,3)
   plot(carr.comp2$ratio_min_max_rate, carr.comp2$TOTALCARBONREPEAT, col='white')
   text(carr.comp2$ratio_min_max_rate, carr.comp2$TOTALCARBONREPEAT, carr.comp2$st)

 #Invasiveness AJE paper--inverse variance weights
   inv1<-read.csv('mcmc_invasive_single_stage.csv')
   inv1$st=as.character(inv1$st)
   inv1$st[inv1$st=='6A/C']<-'6A'
   inv2<-merge(carr.comp, inv1, by='st' , all=TRUE)
    ds<-inv2[,c('ratio_min_max_rate', 'log.inv.prec.age1','log.inv.age1','anaerobic','t.max.deriv1', 'temp','id' )]
   ds<-ds[complete.cases(ds),]
   trans.red<-rgb(1,0,0, alpha=0.1)
    symbols(sqrt(ds$ratio_min_max_rate), ds$log.inv.age1, sqrt(ds$log.inv.prec.age1/pi),inches=0.35,     
            bty='l',fg="white", bg=trans.red,         xlab="Ratio increasing/decreasing Rates", ylab="log(Invasiveness")
   text(sqrt(ds$ratio_min_max_rate), inv2$log.inv.age1, inv2$st, cex=0.75)
  reg.inv<-lme(sqrt(ratio_min_max_rate) ~ log.inv.age1 + anaerobic +t.max.deriv1 + temp , random=~1|id, data=ds )
  round(summary(reg.inv)$tTable,3)
  
 #PS components
   ps1<-read.csv('PS Composition_SS_final.csv')
   ps1$st<-as.character(ps1$Serotype)
   ps2<-merge(carr.comp2, ps1, by='st' , all=TRUE)
   ps2$NAC<-0
   ps2$NAC[ps2$GlcNAc==1 | ps2$GalNAc==1 | ps2$ManNAc==1 |  ps2$ManNAcA==1 |ps2$FucNAc==1 |ps2$PneNAc ==1] <-1
   ps2$uronic<-0
   ps2$uronic[ps2$GalA==1 |ps2$GlcA==1 ]<-1  #GlcA and GalA derived from same pathway
   plot(as.factor(ps2$uronic), ps2$ratio_min_max_rate)
   plot(as.factor(ps2$uronic[ps2$anaerobic=='1']), ps2$ratio_min_max_rate[ps2$anaerobic=='1'])
   plot(as.factor(ps2$uronic[ps2$anaerobic=='2']), ps2$ratio_min_max_rate[ps2$anaerobic=='2'])
   
      ds<-ps2[ps2$anaerobic!='0',c('ratio_min_max_rate', 'uronic','NAC','anaerobic','t.max.deriv1', 'temp','id' )]
   ds<-ds[complete.cases(ds),]
   reg.ps<-lme(ratio_min_max_rate ~ uronic +NAC + anaerobic +t.max.deriv1 + temp + anaerobic*temp, random=~1|id, data=ds )
     round(summary(reg.ps)$tTable,3)
   

```

#Does association between CPS component and growth vary by temp?
In Adrienn's experiments, she found that adding purified CPS had a big effect at 37, no effect or even inhibitory at 33C. Here it appears affect of uronic strongest at 33C with aerobic/catalase growth; slightly weaker at 33C anaerobic. 33C and 37C anaerobic same effect. 37 aerobic weaker effect than 33 aerobic 

```{r effect_temp}

 reg.ps1<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='33' & ds$anaerobic=='1',] )
     round(summary(reg.ps1)$tTable,3)

 reg.ps2<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='33' & ds$anaerobic=='2',] )
      round(summary(reg.ps2)$tTable,3)

  reg.ps3<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='37' & ds$anaerobic=='1',] )
     round(summary(reg.ps3)$tTable,3)

 reg.ps4<-lme(ratio_min_max_rate ~ uronic +NAC + t.max.deriv1   , random=~1|id, data=ds[ds$temp=='37' & ds$anaerobic=='2',] )
     round(summary(reg.ps4)$tTable,3)
     
          #Interaction
reg.ps.int<-lme(ratio_min_max_rate ~ uronic*anaerobic +NAC*anaerobic + anaerobic +t.max.deriv1 + temp + anaerobic*temp, random=~1|id, data=ds )
     round(summary(reg.ps.int)$tTable,3)

```